---
title: 'Aggregated 6 s electricty data - a preliminary study'
author: 'Guy Maskall'
date: '`r format(Sys.Date(), "%d %B, %Y")`'
---

# Introduction

Motivated by a recent [PhD thesis](http://jack-kelly.com/files/writing/jack-kelly-phd-thesis.pdf),
high frequency electricity meter data readings provide a rich source of insight
into domestic consumption. This insight can benefit the consumer through
increased self-awareness, energy suppliers through improved matching between
usage and renewable cost/availability, and distributors through improved
network capacity and preparedness. By being more self-aware of energy use in
real-time, a consumer might think to turn that immersion heater off. By better
understanding the consumer diurnal load profile, a supplier of renewable energy
systems (or electricity) might be able to find best matches with consumers.
And by better anticipating changes in load, a distributor might be able to
power down inefficient baseline (backup) generator for longer.

# Questions

Obvious questions that first come to mind that could be of interest are:

* when is the peak load?
* when is the minimum load?
* when is the load most variable?
* are there any obvious diurnal, weekly, seasonal patterns?

# Data

The aforementioned source [describes](http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/residential/EnergyConsumption/Domestic/UK-DALE-2015/ReadMe_DALE.html#Disaggregated)
some available [data](http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/residential/EnergyConsumption/Domestic/UK-DALE-2015/UK-DALE-disaggregated).
The data originate from five households participating in a test programme.
A richer dataset than usually available from smart meters is available because
a greater granularity of labelled data is available courtesy of additional
meters linked to various power sockets. Here we will only consider the
6 s aggregated data.

```{r, getdata1}
library(tools)
house1_file <- "Data/h1_ch1.dat"
house1_url <- paste0("http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/",
                     "residential/EnergyConsumption/Domestic/UK-DALE-2015/",
                     "UK-DALE-disaggregated/house_1/channel_1.dat")
house1_md5 <- "1e954b140917cb382dedc1e3847fa3ae"

house2_file <- "Data/h2_ch1.dat"
house2_url <- paste0("http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/",
                     "residential/EnergyConsumption/Domestic/UK-DALE-2015/",
                     "UK-DALE-disaggregated/house_2/channel_1.dat")
house2_md5 <- "f779773b5d84261542a8d7c716300eb1"

house3_file <- "Data/h3_ch1.dat"
house3_url <- paste0("http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/",
                     "residential/EnergyConsumption/Domestic/UK-DALE-2015/",
                     "UK-DALE-disaggregated/house_3/channel_1.dat")
house3_md5 <- "dd53383c38b1e0bca126576fb844057d"

house4_file <- "Data/h4_ch1.dat"
house4_url <- paste0("http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/",
                     "residential/EnergyConsumption/Domestic/UK-DALE-2015/",
                     "UK-DALE-disaggregated/house_4/channel_1.dat")
house4_md5 <- "88e47be92f45b2f9d000fe44d8977236"

house5_file <- "Data/h5_ch1.dat"
house5_url <- paste0("http://data.ukedc.rl.ac.uk/simplebrowse/edc/efficiency/",
                     "residential/EnergyConsumption/Domestic/UK-DALE-2015/",
                     "UK-DALE-disaggregated/house_5/channel_1.dat")
house5_md5 <- "424223412a7c702fc365ea9774247fe0"

if (!dir.exists("Data")) {
    dir.create("Data")
}

if (!file.exists(house1_file)) {
    download.file(house1_url, house1_file)
}

if (!file.exists(house2_file)) {
    download.file(house2_url, house2_file)
}

if (!file.exists(house3_file)) {
    download.file(house3_url, house3_file)
}

if (!file.exists(house4_file)) {
    download.file(house4_url, house4_file)
}

if (!file.exists(house5_file)) {
    download.file(house5_url, house5_file)
}

# check md5
check_md5 <- function(filename, md5) {
    if (md5sum(filename) != md5) {
        message(filename, " does not match expected md5sum")
    } else {
        message(filename, " matches expected md5sum: ", md5)
    }
}

check_md5(house1_file, house1_md5)
check_md5(house2_file, house2_md5)
check_md5(house3_file, house3_md5)
check_md5(house4_file, house4_md5)
check_md5(house5_file, house5_md5)
```

## House 1

The data are in whitespace delimited files with Unix timestamp and power
measured.

```{r, house1_data}
library(tidyverse)
h1_data <- read_table(house1_file, col_names = F) %>%
    rename(timestamp = X1, power = X2) %>%
    mutate(timestamp = as.POSIXct(timestamp, origin = "1970/01/01"),
           weekday = weekdays(timestamp),
           time = format(timestamp, "%H:%M:%S"))

h1_data %>% print

h1_data$timestamp %>% range

```

The date range seems to be plausible, suggesting we're interpreting it
correctly.

The time series of the first 5000 rows is shown below.

```{r, house1_first5000_ts, fig.width = 10}
h1_data %>% 
    slice(1:5000) %>% 
    ggplot(aes(x = timestamp, y = power)) + 
    geom_line() +
    labs(title = "House 1 power over time, first 5000 samples")
```

This starts to give us a feel for the time series data from such a system.
We see an interesting initial downwards trend, but mostly the features
comprise periods of fairly level power consumption with step changes
and frequent small ticks. There are some prominent features of relatively
high demand.

### Maximum and minimum load

When was the maximum power load seen for house 1?

```{r, house1_max}
h1_data %>% 
    select(timestamp, power) %>% 
    top_n(1, power)
```

This value and time of day is quite consistent with someone taking a morning
shower.

### Seasonal variation

Now we'll look at standardizing measurement times by reducing the resolution
down to the minutes level.

```{r, h1_standardize}
# 2013 and 2014 are full years in data
h1_data_std <- h1_data %>% 
    #slice(1:5000) %>% 
    filter(timestamp > "2013/01/01", timestamp < "2015/01/01") %>%
    select(timestamp, power) %>% 
    mutate(timestamp = format(timestamp, "%Y-%m-%d %H:%M")) %>% 
    group_by(timestamp) %>% 
    summarise(power_mean = mean(power)) %>% 
    fill(power_mean) %>% #cheap and quick
    mutate(weekday = weekdays(as.POSIXct(timestamp))) %>% 
    separate(timestamp, 
             into = c("year", "month", "day", "hour", "minute"), 
             convert = TRUE) %>%
    mutate(minutes = (hour * 60 + minute)) %>%
    select(year, month, day, weekday, minutes, power_mean) %>%
    spread(minutes, power_mean, fill = 0)

vecs1 <- h1_data_std %>% 
    select(num_range("", 0:1439)) %>% 
    as.matrix()

pca1 <- prcomp(vecs1)

h1_data_std_pca <- h1_data_std %>%
    select(year, month, day, weekday) %>%
    bind_cols(as_data_frame(pca1$x[, 1:3]))
```

```{r, h1_pca, fig.width = 10}
h1_data_std_pca %>% 
    ggplot(aes(x = PC1, y = PC2, colour = weekday)) + 
    geom_point()
```
